---
title: "Categorical Predictors"
subtitle: "Chapter 5 of ALR4, Chapter 14, 15 of LMWR2"
author: "Subrata Paul"
date: "6/3/2020"
output: 
  ioslides_presentation:
    widescreen: true
    hitheme: zenburn
---

<script src="https://ajax.googleapis.com/ajax/libs/jquery/1.12.2/jquery.min.js"></script>

<script>
    $(document).ready(function() {
    $('slide:not(.title-slide, .backdrop, .segue)').append('<footer></footer>');    
    $('footer').attr('url', "https://math5387.web.app");

  })
</script>




<style>
pre {
    display: block;
    font-family: monospace;
    white-space: pre;
    margin: 1em 0px;
    margin-top: 0em;
    margin-right: 0px;
    margin-bottom: -0.2em;
    margin-left: 0px;
}
footer:after {
    font-size: 12pt;
    content: attr(url);
    position: absolute;
    bottom: 5px;
    right: 60px;
    line-height: 1.9;
    display: block;
  }
slides > slide {
  font-family: 'Open Sans', Arial, sans-serif;
  font-size: 26px;
  color: black;
  width: 900px;
  height: 700px;
  margin-left: -450px;
  margin-top: -350px;
  padding: 0px 60px;
  -webkit-border-radius: 5px;
  -moz-border-radius: 5px;
  -ms-border-radius: 5px;
  -o-border-radius: 5px;
  border-radius: 5px;
  -webkit-transition: all 0.6s ease-in-out;
  -moz-transition: all 0.6s ease-in-out;
  -o-transition: all 0.6s ease-in-out;
  transition: all 0.6s ease-in-out;
}
slides > slide > hgroup + article {
  margin-top: 5px;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'center')
```

## Categorical Predictors

A **categorical variable** is a variable that can take on one of a limited, usually fixed, number of possible values, assigning each individual or other unit of observation to a particular group or nominal category on the basis of some qualitative property. 

Categorical predictors are often called **factors**

## Example: UN Data

Consider data from 199 localities (mostly members of the United Nations, but a few areas such as Hong Kong).  The measured variables include:

-	`ppgdp` – the gross national product per person in U.S. dollars
-	`fertility` – average number of children per woman
-	`lifeExpF` – female life expectancy, years
-	`pctUrban` – percentage of population living in urban areas
-	`group` – a factor with level `oecd` for countries that are members of the Organization for Economic Cooperation and Development (OECD) as of May 2012, `africa` for countries on the African continent, and other for all other countries.  No OECD countries are located in Africa.


## Exploring Data for One-Factor Models

If the only predictor is a factor, then the regression model is called a one-factor or one-way design.  

Boxplots of a one-factor model are useful for comparing different levels of each factor.

-	The thick middle line indicates the median value.
-	The box extends to the 25th and 75th percentiles.  The distance between the quartiles is known as the **interquartile range (IQR)**.
-	The **whiskers** generally extend to the most extreme values that are within 1.5 IQRs from the **quartiles**.
-	Values outside the whiskers are outliers and are indicated by a dot or star.

## Boxplot Example


```{r, fig.align="center"}
data(UN11, package = 'alr4')
boxplot(lifeExpF ~ group, data = UN11)
```

## More Details 
```{r, message=F}
library(ggstatsplot)
UN11$country = row.names(UN11)
ggbetweenstats(data = UN11, plot.type = 'box',x = group, y = lifeExpF, outlier.tagging = T, outlier.label =  country)
```

## What we get?

-The oecd countries generally have the largest female life expectancy, with africa having the lowest.  

- Nauru and Afghanistan have unusually low female life expectancy for the other group.  

- Japan has high female life expectancy, and Turkey has low life expectancy, compared to the other members of the oecd.  

- The variation of female life expectancy in oecd is smallest, but in africa is largest.  


## Defining Factors as Regressors


## Defining Factors as Regressors
**Dummy** or **indicator** variables are used to include categorical predictors in a regression model.

A dummy variable $U_j$ for factor level $j$ is 1 if an observation has level $j$, but 0 otherwise.

- -1 and 1, or 1 and 2, are sometimes used, but this is less common and more difficult to interpret.
-	Assignment labels are mostly arbitrary and do not affect the results.

Since group has $d=3$ levels, the $j$th dummy variable $U_j$ for the factor $j=1,2,…,d$ has $i$th value $u_ij$, for $i=1,2,…,n$, given by

$$u_{ij} = \begin{cases}1 & \text{ if group}_i = j\text{th category of group}\\ 0 &\text{ otherwise}\end{cases}$$

## Try in `R`

```{r}
levels(UN11$group)
U1 = with(UN11, (group == levels(group)[1])+0)
U2 = with(UN11, (group == levels(group)[2])+0)
U3 = with(UN11, (group == levels(group)[3])+0)
head(data.frame(group = UN11$group, U1, U2, U3))
```

U_1 is the dummy variable for oecd, U_2 is the dummy variable for other, and U_3 for africa.

## One-Factor Model

Regression coefficients are generally called effects in this setting.  

How can we build a model using the form y=X\beta+ϵ?

What would happen if we fit the model

$E(lifeExpF│group)=\beta_0+\beta_1 U_1+\beta_2 U_2+\beta_3 U_3?$

Why?

## This is why!

In that case, $X$ would be rank deficient since $U_1+U_2+U_3=1$, which will always match the intercept.
	
- The columns of $X$ would be linearly dependent.

We only need $d-1$ dummy variables to represent $d$ levels because the last level represented when all the other dummy variables were 0.

-	Any of the $d$ dummy variables could be excluded, though generally it is the first or last level.
-	`R` drops the first level by default.
-	The level dropped is known as the **reference** or **baseline** level.
-	In `R`, this method for creating the dummy variables is known as the **treatment contrast**.

## A different solution

Other possibilities for avoiding linear dependence are:

-	Drop $\beta_0$ from the model.
-	Assume $\sum_{i=1}^{p-1}\beta_i =0$.
  - This is known as a **sum contrast** and does NOT use dummy variables.
  - Interpretation is generally more difficult.

## Models

Using a treatment contrast with the first level of `group` as the reference level (`oecd`), our model becomes
$$E(lifeExpF│group)=\beta_0+\beta_2 U_2+\beta_3 U_3.$$

Since `group = oecd` implies $U_2=U_3=0$,
$$E(lifeExpF│group=oecd)=\beta_0+\beta_2 0+\beta_3 0=\beta_0.$$

Since `group = other` implies $U_2=1 \text{ and } U_3=0$,
$$E(lifeExpF│group=other)=\beta_0+\beta_2 1+\beta_3 0=\beta_0+\beta_2.$$

Since `group = africa` implies $U_2=0 \text{ and } U_3=1$,
$$E(lifeExpF│group=africa)=\beta_0+\beta_2 0+\beta_3 1=\beta_0+\beta_3.$$

## Fit in `R`

* Method 1

```{r}
lm(lifeExpF ~ U2 + U3, data = UN11)
```

* Method 2

```{r}
lm(lifeExpF ~ group, data = UN11)
```

## `LifeExpF` of different groups

- $\hat{E}(Lifeexpf│group=oecd)=\hat{\beta}_0=82.45.$
- $\hat{E}(Lifeexpf│group=other)=\hat{\beta}_0+\hat{\beta}_2=82.45-7.12=75.33.$
- $\hat{E}(Lifeexpf│group=africa)=\hat{\beta}_0+\hat{\beta}_3=82.45-22.67=59.79.$


## Interpretation 

```{r}
lm(lifeExpF ~ group, data = UN11)
```


## Interpretation 

```{r}
lm(lifeExpF ~ group, data = UN11)
```

-	The expected female life expectancy for OECD nations is 82.45 years.
-	The expected female life expectancy for other nations is 7.12 years less than nations in OECD.
-	The expected female life expectancy for African nations is 22.67 years less than nations in OECD.

## Interesting Facts

-	$\hat{\beta}_0$ is simply the sample mean of the responses in the oecd group.
-	$\hat{\beta}_2$ is the difference between the sample mean of the responses for the other group and the oecd group.
-	$\hat{\beta}_3$ is the difference between the sample mean of the responses for the africa group and the oecd group.

```{r}
with(UN11, tapply(lifeExpF, group, mean))
```


## Effect Plot

```{r}
library(effects)
lmod1 = lm(lifeExpF ~ group, data = UN11)
plot(predictorEffect('group',lmod1))
```


## Pairwise comparison

```{r}
library(emmeans)
group.emm.s = emmeans(lmod1, specs = 'group')
knitr::kable(pairs(group.emm.s))
```


## Factors and Quantitative Predictors

It is common to study the effect of a factor AFTER adjusting for one or more other quantitative regressors.  

If we have a mixture of factors and quantitative variables in our data, it’s useful to create a scatterplot of the response versus the covariate that distinguishes the observations for each level.

-	This helps us assess whether the relationship between the response and quantitative regressor differs for the levels of the factor.


## Scatter Plot 

```{r}
library(ggplot2)
ggplot(data = UN11, aes(x = ppgdp, y = lifeExpF, col = group, shape = group))+
  geom_point()
```

## Helpful One 
```{r}
ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, col = group, shape = group))+
  geom_point()
```

## What we see?

-	The linear relationship between lifeExpF and log(ppgdp) is fairly weak for the africa locations
-	The linear relationship is reasonably strong for the other locations
-	The linear relationship is reasonably strong for oecd locations.
-	It’s unclear whether the average rate of change between lifeExpF and log(ppgdp) (the slope) is the same for the three factor levels.
-	The average lifeExpF seems to differ vertically for the same levels of log(ppgdp) for the different factor levels.

## Different Models to Consider 

Suppose we have a response $y$, a quantitative regressor $x$, and a two-level factor variable represented by a dummy variable $u$:

$$u = \begin{cases}0 & \text{ for the reference level}\\ 1 & \text{ for the treatment level.}\end{cases}$$

Consider several possible regression models:

-	The same regression line for both levels, y=β_0+β_1 x+ϵ.
    -	In R: `y ~ x`
-	A factor predictor but no quantitative predictor, y=β_0+β_2 u+ϵ.
    -	In R: `y ~ u`
    -	This is a **one-way model**.
  
## Different Models to Consider 

-	Separate regression lines for each group having the same slope, $y=\beta_0+\beta_1 x+\beta_2 u+\epsilon$.
    -	In R: `y ~ x + u.`
    -	This is known as a parallel lines or main effects model.
    -	$\beta_2$ represents the vertical distance between the regression lines (the effect of the treatment).

-	Separate lines for each group with different slopes, $y=\beta_0+\beta_1 x+\beta_2 u+\beta_3 xu+\epsilon$.
    -	In R: `y ~ x + u + x:u` or `y ~ x*u`
	  - This is known as a separate lines or interaction model.
	  - `x:u` means the interaction between $x$ and $u$.
	  - `x*u` means the cross between $x$ and $u$ ($x$, $u$, and the interaction between $x$ and $u$). 

## Example

```{r}
lmodi = lm(lifeExpF ~ group*log(ppgdp), data = UN11)
head(model.matrix(lmodi))
```



## Example
```{r, results='asis', message=F}
library(knitr)
kable(summary(lmodi)$coefficients)
```

## Example
```{r}
lm(lifeExpF ~ log(ppgdp), data = UN11[UN11$group=='oecd',])$coefficients
lm(lifeExpF ~ log(ppgdp), data = UN11[UN11$group=='other',])$coefficients
lm(lifeExpF ~ log(ppgdp), data = UN11[UN11$group=='africa',])$coefficients
```

## Example

$$\hat{E}(lifeExpF│ppgdp=x,group=oecd)= $$





$$\hat{E}(lifeExpF│ppgdp=x,group=other)= $$





$$\hat{E}(lifeExpF│ppgdp=x,group=africa)= $$

## Example
```{r}
ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, col = group, shape = group))+
  geom_point() + geom_smooth(method = 'lm')
```


## Effect Plot

```{r}
par(mfrow = c(1,2))
plot(predictorEffect('ppgdp',lmodi), lines = list(multiline = T))
```


## Effect Plot with Transformation

```{r}
UN11$logppgdp = log(UN11$ppgdp)
plot(predictorEffect('logppgdp',lm(lifeExpF ~ group*logppgdp, data = UN11)), lines = list(multiline = T))
```


## Main Effect Model 
Examining the fit of the interaction model to the data suggests that the slope might be the same for all groups. 

A **main effects model** assumes the slope is the same for all factor levels but allows each group to have its own intercept.  

-	Main effects models are easier to interpret since the effect of the quantitative regressor is the same for all levels of the factor.
-	This model is called the **Analysis of Covariance (ANCOVA)** when the factor levels indicate a randomly assigned treatment for subjects.




## Example 

```{r}
lmodm = lm(lifeExpF ~ log(ppgdp) + group, UN11)
kable(summary(lmodm)$coefficients)
```


## Plot

```{r}
ggplot(data = UN11, aes(x = log(ppgdp), y = lifeExpF, col = group, shape = group))+
  geom_point() + geom_smooth(method = 'lm', mapping= aes(y = predict(lmodm, UN11)))
```