---
title: "Homework 6 Key"
author: "Subrata Paul"
date: "10/25/2020"
output: pdf_document
---

```{r echo=F}
library(xtable)
options(scipen=999)
options(xtable.comment = FALSE)
```

## Problem 3.3

(a) The predictors sex and income are significant at the 0.05 level. 
```{r message=F, results='asis'}
lmod = lm(gamble ~., data = faraway::teengamb)
xtable(lmod)
```

(b) We predict that females will spend about 22 lbs less per year on gambling than their male counterparts, assuming that their socioeconomic status, income, and verbal scores are the same. 

(c)

$H_0:\ \beta_{\text{sex}} = \beta_{\text{status}} = \beta_{\text{verbal}}=0|\beta_{\text{income}}\neq 0$

$H_a:\ \beta_{\text{sex}} \text{ or } \beta_{\text{status}} \text{ or } \beta_{\text{verbal}}\neq 0|\beta_{\text{income}}\neq 0$

```{r}
full_mod = lm(gamble ~ ., data = faraway::teengamb)
reduced_mod = lm(gamble ~ income, data = faraway::teengamb)
(rssf = deviance(full_mod))
(rssr = deviance(reduced_mod))
(f = ((rssr - rssf)/3)/(rssf/42))
1 - pf(f, df1 = 3, df2 = 42)
```


$$F = \frac{(`r round(rssr,2)` - `r round(rssf,2)`)/ `r reduced_mod$df.residual - full_mod$df.residual`}{`r round(rssf,2)`/`r full_mod$df.residual`} = `r f`$$

p-value = $P(F_{3,42} \geq 4.13)$ = 0.012

There is moderate evidence that at least one of the predictors sex, status, or verbal should be included in the regression model of gamble that already contains the income predictor.

## Problem 3.4

(a)

$H_0:\ \beta_{\text{salary}} = 0| \beta_{\text{ratio}}\neq 0, \beta_{\text{expend}}\neq 0$

$H_0:\ \beta_{\text{salary}} \neq 0| \beta_{\text{ratio}}\neq 0, \beta_{\text{expend}}\neq 0$

```{r}
fmod = lm(total ~ expend + ratio + salary, data = faraway::sat)
summary(fmod)
```

$$t = \frac{`r fmod$coefficients['salary']`}{`r summary(fmod)$coefficients[4,2]`} = `r summary(fmod)$coefficients[4,3]` $$

p-value = $2P(T_{46} \geq `r abs(summary(fmod)$coefficients[4,3])`) = `r summary(fmod)$coefficients[4,4]`$

There is weak evidence to conclude the salary predictor should be included in the model of sat scores that already has the ratio and expend predictors. 

$H_0:\ \beta_{\text{salary}} = \beta_{\text{ratio}} = \beta_{\text{expend}}= 0$

$H_0:\ \beta_{\text{salary}} \neq 0 \text{ or } \beta_{\text{ratio}} \neq 0 \text{ or } \beta_{\text{expend}}\neq  0$

```{r}
rmod = lm(total ~ 1, faraway::sat)
(rssf = deviance(fmod))
(rssr = deviance(rmod))
(f = ((rssr - rssf)/(rmod$df.residual - fmod$df.residual))/(rssf/fmod$df.residual))
1 - pf(f, df1 = rmod$df.residual - fmod$df.residual, df2 = fmod$df.residual)
```



$$F = \frac{(`r round(rssr,2)` - `r round(rssf,2)`)/ `r rmod$df.residual - fmod$df.residual`}{`r round(rssf,2)`/`r fmod$df.residual`} = `r f`$$

p-value = $P(F_{3,42} \geq 4.13)$ = `r 1 - pf(f, df1 = rmod$df.residual - fmod$df.residual, df2 = fmod$df.residual)`.

There is moderate evidence to conclude at least one of the predictors salary, ratio, or expend should be used in modeling the sat predictor.


(b)

F test: 

$H_0:\ \beta_\text{takers}=0|\beta_\text{ratio}\neq 0, \beta_\text{expend}\neq 0 , \beta_\text{salary}\neq 0$

$H_a:\ \beta_\text{takers}\neq 0|\beta_\text{ratio}\neq 0, \beta_\text{expend}\neq 0 , \beta_\text{salary}\neq 0$

```{r}
options(scipen=5)
rmod = lm(total ~ expend + ratio + salary, faraway::sat)
fmod = lm(total ~ expend + ratio + salary + takers, faraway::sat)
(rssf = deviance(fmod))
(rssr = deviance(rmod))
(f = ((rssr - rssf)/(rmod$df.residual - fmod$df.residual))/(rssf/fmod$df.residual))
1 - pf(f, df1 = rmod$df.residual - fmod$df.residual, df2 = fmod$df.residual)
```



$$F = \frac{(`r round(rssr,2)` - `r round(rssf,2)`)/ `r rmod$df.residual - fmod$df.residual`}{`r round(rssf,2)`/`r fmod$df.residual`} = `r f`$$

p-value = $P(F_{3,42} \geq 4.13)$ = `r 1 - pf(f, df1 = rmod$df.residual - fmod$df.residual, df2 = fmod$df.residual)`.

There is very strong evidence to conclude that the takers predictor should be included in the model of sat scores that already has the ratio, salary, and expend predictors. 

t test:

```{r results='asis'}
library(xtable)
xtable(fmod)
```

$$t = \frac{`r fmod$coefficients['takers']`}{`r summary(fmod)$coefficients[5,2]`} = `r summary(fmod)$coefficients[5,3]` $$

p-value = $2P(T_{46} \geq `r abs(summary(fmod)$coefficients[5,3])`) = `r summary(fmod)$coefficients[5,4]`$


Notice that $t^2 = F = 157.74$


